---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2025"
subtitle: "Assignment 2 - Due date 01/27/26"
author: "Lauren Xu"
output: pdf_document
geometry: margin=2.54cm
---

## Submission Instructions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github.   

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A02_Sp26.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

## R packages

R packages needed for this assignment:"forecast","tseries", and "dplyr". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package here
library(readxl)
library(forecast)
library(tseries)
library(dplyr)
library(ggplot2)
```


## Data set information  

Consider the data provided in the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx" on our **Data** folder. The data comes from the US Energy Information and Administration and corresponds to the December 2025 Monthly Energy Review. The spreadsheet is ready to be used. Refer to the file "M2_ImportingData_XLSX.Rmd" in our Lessons folder for instructions on how to read $.xlsx$ files.

```{r}
#Importing data set
getwd()
energy_data1 <- read_excel(path="/Users/xxx/R/Time Series_R/TSA/Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",skip = 12, sheet="Monthly Data",col_names=FALSE) 
```


## Question 1
You will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. Use the command head() to verify your data.

```{r}
read_col_names <- read_excel(path="/Users/xxx/R/Time Series_R/TSA/Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx",skip = 10,n_max = 1, sheet="Monthly Data",col_names=FALSE) 

colnames(energy_data1) <- read_col_names

head(energy_data1)

ts_energy_data1 <- energy_data1[,4:6]
```


## Question 2
Transform your data frame in a time series object and specify the starting point and frequency of the time series using the function ts().

```{r include=FALSE}
ts_energy_data1 <- ts(ts_energy_data1, start = c(1973, 1),frequency = 12)

ts_energy_data1
```


## Question 3
Compute mean and standard deviation for these three series.

```{r}
ts.mean <- colMeans(ts_energy_data1, na.rm = TRUE)
ts.sd <- apply(ts_energy_data1, 2, sd, na.rm = TRUE)

ts.mean
ts.sd
```


## Question 4
Display and interpret the time series plot for each of these variables. Try to make your plot as informative as possible by writing titles, labels, etc. For each plot add a horizontal line at the mean of each series in a different color.

```{r}
mean_biomass <- mean(ts_energy_data1[,1], na.rm = TRUE)
mean_renewable <- mean(ts_energy_data1[,2], na.rm = TRUE)
mean_hydroelectric <- mean(ts_energy_data1[,3], na.rm = TRUE)


autoplot(ts_energy_data1[,1]) +
  geom_hline(
    aes(yintercept = mean_biomass, color = "Mean")
  )+
  scale_color_manual(values = c("Mean" = "blue"), name = "") +
  xlab("Time") +
  ylab("Total Biomass Energy Consumption (Trillion Btu)") +
  ggtitle("Monthly Biomass Energy Consumption with Mean Line")


autoplot(ts_energy_data1[,2]) +
  geom_hline(
  aes(yintercept = mean_renewable, color = "Mean")
  )+
  scale_color_manual(values = c("Mean" = "purple"), name = "") +
  xlab("Time") +
  ylab("Total Renewable Energy Consumption (Trillion Btu)") +
  ggtitle("Monthly Renewable Energy Consumption with Mean Line")


autoplot(ts_energy_data1[,3]) +
  geom_hline(
    aes(yintercept = mean_hydroelectric, color = "Mean")
  )+
  scale_color_manual(values = c("Mean" = "red"), name = "") +
  xlab("Time") +
  ylab("Total Hydroelectric Power Consumption (Trillion Btu)") +
  ggtitle("Monthly Hydroelectric Power Consumption with Mean Line")
```


## Question 5
Compute the correlation between these three series. Are they significantly correlated? Explain your answer.

```{r}
cor.test(ts_energy_data1[,1],
         ts_energy_data1[,2])
cor.test(ts_energy_data1[,1],
         ts_energy_data1[,3])
cor.test(ts_energy_data1[,2],
         ts_energy_data1[,3])
```
1. Renewable energy and Biomass energy are strongly and significantly correlated, since correlation is 0.965 and p-value<0.05. 
2. Biomass is negatively and not strongly correlated with the Hydroelectric power, as the correlations are -0.135. The relationship is statistically significant (p-value<0.05).
3. Renewable energy is also negatively and not strongly correlated (-0.058) with hydroelectric power and the relationship is not significant (p-value>0.05).

## Question 6
Compute the autocorrelation function from lag 1 up to lag 40 for these three variables. What can you say about these plots? Do the three of them have the same behavior?

```{r}
acf_biomass <- acf(ts_energy_data1[,1], lag.max = 40, plot = TRUE, main = ("ACF of Total Biomass Energy Consumption"))
acf_renewable <- acf(ts_energy_data1[,2], lag.max = 40, plot = TRUE, main = ("ACF of Total Renewable Energy Consumption"))
acf_hydroelectric <- acf(ts_energy_data1[,3], lag.max = 40, plot = TRUE, main = ("ACF of Hydroelectric Power Consumption"))
```
Their ACFs decrease as lags increase.

Biomass: Extremely persistent with very slow decay.
Renewable: Very similar to the renewable, strong persistence.
Hydroelectric: Moderate persistent and autocorrelation declines faster.


## Question 7
Compute the partial autocorrelation function from lag 1 to lag 40 for these three variables. How these plots differ from the ones in Q6?

```{r}
pacf_biomass <- pacf(ts_energy_data1[,1], lag.max = 40, plot = TRUE, main = ("PACF of Total Biomass Energy Consumption"))
pacf_renewable <- pacf(ts_energy_data1[,2], lag.max = 40, plot = TRUE, main = ("PACF of Total Renewable Energy Consumption"))
pacf_hydroelectric <- pacf(ts_energy_data1[,3], lag.max = 40, plot = TRUE, main = ("PACF of Hydroelectric Power Consumption"))
```
In ACF, autocorrelations remain large for many lags, especially in Renewable and Biomass plots. In PACF, however, Lag 1 is very large and significant while most other lags are small and insignificant.

